---
title: "Scenario simulator"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4
)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidytargets)
library(crew)
```

Signature gene ranks are drawn as **distinct** positions in `1:gene_size`, because
real ranks are a permutation: two genes cannot occupy the same slot. The beta
shape sets where the signature prefers to sit, and sampling without replacement
enforces that only `gene_size - k` slots exist above position `k`.

The shape family is chosen so each scenario has its mode on the boundary and a
monotone density, which keeps the benchmark clean:

| scenario | weights | mode |
|---|---|---|
| UP | `Beta(shape, 1)` | at 1 |
| DOWN | `Beta(1, shape)` | at 0 |
| flat | `Beta(1, 1)` | none (uniform) |

A mean pinned near 1 is not available here: a 1000-gene signature packed into the
top 1000 slots already has mean 0.95, so asking for 0.98 forces the sampler to
spill downward and produces a hump instead of a boundary mode.

The rank cap is a proxy for sparsity. Genes below it are tied at the bottom and
carry no information, so they score 0. A bigger cell lowers the cap and lets more
signature genes clear it; the score must not drift upward for that reason alone.
The test is therefore on the **gradient** of score against `rank_cap`, which
should be flat.

## Functions

```{r functions}
# responder_fraction < 1 mixes responders with genes that sit at background,
# which is the realistic case of a signature where only part of it reacts.
simulate_peaks <- function(direction = "flat",
                           shape = 5,
                           responder_fraction = 1,
                           signature_size = 1000L,
                           gene_size = 10000L) {
  shapes <- switch(direction,
    up   = c(shape, 1),
    down = c(1, shape),
    flat = c(1, 1)
  )
  slots <- seq_len(gene_size)
  responders <- stats::dbeta((slots - 0.5) / gene_size, shapes[1], shapes[2])
  weights <- responder_fraction * responders + (1 - responder_fraction)
  sort(sample(slots, signature_size, replace = FALSE, prob = weights))
}

# Exponential tilt f(x) ~ exp(theta * x) on [0, 1]: one unconstrained parameter,
# UP for theta > 0 and DOWN for theta < 0, meeting smoothly at the uniform. The
# Beta parameterisation would kink at 0 and break the curvature estimate there.
# Z(theta) = (e^theta - 1)/theta is positive on both sides, so log it on
# magnitudes; the small-theta branches are the series expansions.
tilt <- list(
  logZ = function(theta) {
    if (abs(theta) < 1e-6) log1p(theta / 2 + theta^2 / 6)
    else log(abs(expm1(theta))) - log(abs(theta))
  },
  mean = function(theta) {
    if (abs(theta) < 1e-6) 0.5 + theta / 12 else 1 / (1 - exp(-theta)) - 1 / theta
  },
  cdf = function(q, theta) {
    if (abs(theta) < 1e-6) q else expm1(theta * q) / expm1(theta)
  },
  # E[x | x <= q], the value the fit imputes for every censored gene.
  cond_mean = function(q, theta) {
    if (abs(theta) < 1e-6) q / 2 else q / (1 - exp(-theta * q)) - 1 / theta
  }
)

# Four ways to turn one (peaks, rank_cap) pair into a score, all on 0-1, plus
# diagnostics for the last one. local() puts `tilt` in the function's enclosing
# environment so it travels with the function to a parallel worker.
score_variants <- local({
  tilt <- tilt
  function(peaks, rank_cap, gene_size = 10000L) {
  detected <- peaks > rank_cap
  effective_size <- sum(detected)
  signature_size <- length(peaks)
  censored_size <- signature_size - effective_size
  relative <- sum(peaks[detected] - rank_cap) / (gene_size - rank_cap)

  # Tie-aware Mann-Whitney against a fixed background of all gene_size genes.
  # A detected gene outranks every gene below it; a censored gene sits at 0 and
  # ties with the rank_cap other censored genes, earning half credit on each.
  wins <- sum(peaks[detected] - 1)
  ties <- censored_size * rank_cap / 2

  # Censored MLE: detected genes contribute their exact position, censored genes
  # contribute only the probability of falling below the cap. Nothing is
  # rescaled, so the cap enters the likelihood rather than the estimate.
  positions <- (peaks - 0.5) / gene_size
  cap_quantile <- rank_cap / gene_size
  # The data enter only through these three numbers, so each evaluation of the
  # likelihood is O(1) rather than O(signature_size).
  position_sum <- sum(positions[detected])
  negative_loglik <- function(theta) {
    -(theta * position_sum - effective_size * tilt$logZ(theta) +
      censored_size * log(max(tilt$cdf(cap_quantile, theta), 1e-300)))
  }
  theta <- stats::optimize(negative_loglik, c(-60, 60), tol = 1e-8)$minimum

  # Observed information by central difference, then the delta method onto the
  # 0-1 mean scale. Once the signature is fully censored the likelihood flattens
  # and this blows up, which is the estimator declaring itself undetermined.
  step <- 1e-3
  curvature <- (negative_loglik(theta + step) - 2 * negative_loglik(theta) +
                negative_loglik(theta - step)) / step^2
  slope <- (tilt$mean(theta + step) - tilt$mean(theta - step)) / (2 * step)

  c(
    over_total     = relative / signature_size,
    over_effective = if (effective_size > 0) relative / effective_size else 0,
    auc            = (wins + ties) / (signature_size * (gene_size - 1)),
    mle            = tilt$mean(theta),
    mle_se         = abs(slope) / sqrt(max(curvature, 1e-12)),
    mle_theta      = theta,
    # Worst case imputation error, assuming nothing at all about the true shape:
    # the score and the truth differ only in a conditional mean that must lie in
    # [0, cap], so the gap cannot exceed the censored fraction times the cap.
    bias_bound     = (censored_size / signature_size) * cap_quantile
  )
}})

# One signature per replicate, scored at every cap: the cap effect is then
# within-signature rather than confounded with draw-to-draw noise.
score_one <- function(settings, rank_caps, simulate_peaks, score_variants) {
  peaks <- simulate_peaks(
    direction = settings$direction,
    responder_fraction = settings$responder_fraction
  )
  scored <- vapply(rank_caps, function(cap) score_variants(peaks, cap), numeric(7))
  data.frame(
    scenario = settings$scenario,
    rep = settings$rep,
    rank_cap = rank_caps,
    t(scored)
  )
}

# Scale-free readout: how much of the score's spread the cap explains, and how
# far the fitted curve travels across the swept range relative to its own mean.
methods_all <- c("mle", "auc", "over_effective", "over_total")

# Bias and dispersion are reported separately: an estimator that keeps its mean
# and widens honestly is preferable to one that drifts while looking confident.
gradient_table <- function(sweep, scenario, methods = methods_all) {
  dat <- sweep[sweep$scenario == scenario, ]
  caps <- sort(unique(dat$rank_cap))
  do.call(rbind, lapply(methods, function(method) {
    y <- dat[[method]]
    x <- dat$rank_cap
    model <- stats::lm(y ~ x + I(x^2))
    fitted_curve <- stats::predict(model, data.frame(x = caps))
    by_cap <- tapply(y, x, stats::sd)
    data.frame(
      method = method,
      mean_score = mean(y),
      drift = max(fitted_curve) - min(fitted_curve),
      relative_drift = (max(fitted_curve) - min(fitted_curve)) / max(abs(mean(y)), 1e-9),
      sd_low_cap = by_cap[[1]],
      sd_high_cap = by_cap[[length(by_cap)]],
      sd_ratio = by_cap[[length(by_cap)]] / max(by_cap[[1]], 1e-9)
    )
  }))
}

method_colours <- c(over_total = "#BBBBBB", over_effective = "#C0392B",
                    auc = "#2E86C1", mle = "#1E8449")

to_long <- function(dat) {
  bind_rows(lapply(methods_all, function(m) {
    data.frame(rank_cap = dat$rank_cap, score = dat[[m]], method = m)
  }))
}

plot_invariance <- function(sweep, scenario) {
  long <- to_long(sweep[sweep$scenario == scenario, ])
  ggplot(long, aes(x = rank_cap, y = score, colour = method)) +
    geom_point(alpha = 0.2, size = 0.8) +
    geom_smooth(method = "lm", formula = y ~ x + I(x^2), se = FALSE, linewidth = 0.7) +
    geom_hline(yintercept = 0.5, linetype = "dotted", colour = "grey40") +
    scale_colour_manual(values = method_colours) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = "rank_cap", y = "score", colour = NULL, subtitle = scenario) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# An SE is only worth reporting if it matches the spread actually observed across
# replicates. The ratio should sit near 1 while the estimate is usable, and run
# away once the signature is fully censored.
se_calibration <- function(sweep, scenario) {
  dat <- sweep[sweep$scenario == scenario, ]
  dat |>
    group_by(rank_cap) |>
    summarise(
      mle_mean = mean(mle),
      empirical_sd = sd(mle),
      model_se = stats::median(mle_se),
      .groups = "drop"
    ) |>
    mutate(se_over_sd = model_se / empirical_sd) |>
    filter(rank_cap %in% seq(500, 9500, by = 1000)) |>
    as.data.frame()
}

plot_se_calibration <- function(sweep, scenario) {
  dat <- sweep[sweep$scenario == scenario, ]
  spread <- dat |>
    group_by(rank_cap) |>
    summarise(mle = mean(mle), lower = mean(mle) - stats::median(mle_se),
              upper = mean(mle) + stats::median(mle_se), .groups = "drop")
  ggplot(spread, aes(x = rank_cap, y = mle)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#1E8449", alpha = 0.25) +
    geom_line(colour = "#1E8449", linewidth = 0.8) +
    geom_hline(yintercept = 0.5, linetype = "dotted", colour = "grey40") +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = "rank_cap", y = "mle +/- model SE", subtitle = scenario) +
    theme_minimal()
}

# The complement of the invariance plot: does the spread widen as information
# disappears, or does the estimator collapse onto a constant and look certain?
plot_dispersion <- function(sweep, scenario) {
  long <- to_long(sweep[sweep$scenario == scenario, ])
  spread <- long |>
    group_by(method, rank_cap) |>
    summarise(sd = sd(score), .groups = "drop")
  ggplot(spread, aes(x = rank_cap, y = sd, colour = method)) +
    geom_line(linewidth = 0.7) +
    scale_colour_manual(values = method_colours) +
    labs(x = "rank_cap", y = "SD across replicates", colour = NULL, subtitle = scenario) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Core plot: density over a barcode, with the cap and the scores annotated.
plot_peaks <- function(peaks, rank_cap, gene_size = 10000L, title = NULL) {
  df <- data.frame(peak = peaks)
  detected <- peaks > rank_cap
  scores <- score_variants(peaks, rank_cap, gene_size)
  label <- paste0(
    "effective_signature_size = ", sum(detected),
    "\nmle = ", round(scores[["mle"]], 3), " +/- ", signif(scores[["mle_se"]], 2),
    "\nauc = ", round(scores[["auc"]], 3),
    "\nover_effective = ", round(scores[["over_effective"]], 3)
  )

  p_density <- ggplot(df, aes(x = peak)) +
    geom_density(fill = "#4C97CB", alpha = 0.6, colour = NA) +
    geom_vline(xintercept = rank_cap, colour = "red", linewidth = 0.8) +
    annotate("text", x = gene_size, y = Inf, label = label,
             hjust = 1.05, vjust = 1.2, size = 3.5, colour = "red") +
    scale_x_continuous(limits = c(1, gene_size), expand = c(0, 0)) +
    labs(x = NULL, y = "Density", title = title) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())

  p_barcode <- ggplot(df, aes(x = peak, y = 0)) +
    geom_segment(aes(xend = peak, yend = 1), linewidth = 0.35, colour = "#333333") +
    scale_x_continuous(limits = c(1, gene_size), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    labs(x = "Peak", y = NULL) +
    theme_void() +
    theme(axis.text.x = element_text(), axis.title.x = element_text())

  p_density / p_barcode + plot_layout(heights = c(2, 1))
}

plot_peak_samples <- function(direction,
                              rank_cap,
                              responder_fraction = 1,
                              gene_size = 10000L,
                              signature_size = 1000L) {
  plot_peaks(
    simulate_peaks(
      direction = direction,
      responder_fraction = responder_fraction,
      signature_size = signature_size,
      gene_size = gene_size
    ),
    rank_cap = rank_cap,
    gene_size = gene_size
  )
}

plot_awkward_shape <- function(shape_name, rank_cap, gene_size = 10000L, seed = 31) {
  set.seed(seed)
  peaks <- awkward_shapes[[shape_name]]()
  plot_peaks(peaks, rank_cap, gene_size,
             title = paste0(shape_name, "  |  rank_cap = ", rank_cap))
}

# Signatures the tilt family cannot represent, for the assumptions section.
awkward_shapes <- list(
  "monotone UP (not clustered)" = function() {
    slots <- seq_len(10000L)
    sort(sample(slots, 1000L, prob = stats::dbeta((slots - 0.5) / 10000, 5, 1)))
  },
  "band 5000-7000"          = function() sort(sample(5001:7000, 1000L)),
  "band 2000-4000"          = function() sort(sample(2001:4000, 1000L)),
  "narrow block at 6000"    = function() sort(sample(5501:6500, 1000L)),
  "two clusters, 2k and 9k" = function() sort(c(sample(1500:2500, 500L),
                                                sample(8500:9500, 500L)))
)

# Does the detected part of the signature actually look like the fitted tilt?
# A standard error cannot see model misspecification, but this can.
tilt_gof <- function(peaks, rank_cap, theta, gene_size = 10000L) {
  above <- sort((peaks[peaks > rank_cap] - 0.5) / gene_size)
  # With nothing detected there is no fitted shape left to test.
  if (length(above) < 5) return(NA_real_)
  below_cap <- tilt$cdf(rank_cap / gene_size, theta)
  expected <- (vapply(above, tilt$cdf, numeric(1), theta = theta) - below_cap) /
    (1 - below_cap)
  max(abs(expected - (seq_along(above) - 0.5) / length(above)))
}

shape_table <- function(shape_name, caps, gene_size = 10000L, seed = 31) {
  set.seed(seed)
  peaks <- awkward_shapes[[shape_name]]()
  truth <- mean((peaks - 0.5) / gene_size)
  do.call(rbind, lapply(caps, function(cap) {
    scored <- score_variants(peaks, cap, gene_size)
    data.frame(
      cap = cap,
      detected = sum(peaks > cap),
      truth = truth,
      mle = scored[["mle"]],
      bias = scored[["mle"]] - truth,
      bound = scored[["bias_bound"]],
      se = scored[["mle_se"]],
      ks = tilt_gof(peaks, cap, scored[["mle_theta"]], gene_size)
    )
  }))
}
```

## The four variants

Two things must be judged separately. An estimator should keep its **mean** as the
cap moves, and its **spread** should widen as information disappears. Drifting is
a bias; narrowing while drifting is worse, because the score then looks confident
about the wrong answer. Each scenario below therefore gets both an invariance plot
and a dispersion plot.

`over_effective` rescales the surviving ranks onto `(cap, G]`, so it asks where the
detected genes sit *within the observable window*. It is **exactly** unbiased for
the DOWN and flat scenarios, and not by luck: for `f(x) ~ (1-x)^(a-1)`, setting
`u = (x - c)/(1 - c)` gives `1 - x = (1 - c)(1 - u)`, so the truncated density is
`~ (1 - u)^(a-1)`, the same family with the same parameter, and `E[u] = 1/(1 + a)`
at every cap. The UP family `x^(a-1)` is *not* closed under that operation: it
flattens toward uniform, so the score has to decay to 0.5. That decay is the
normaliser, not the biology. `over_total` inherits the same artefact and decays
with the detected fraction on top of it.

`auc` compares every gene against a fixed background of all `G` genes, so nothing
is rescaled when the cap moves. It is algebraically the same statistic as imputing
censored genes at `cap / 2` (the two agree to 1e-5), but the Mann-Whitney form
makes the tie convention explicit. It holds its mean for UP. Its weakness is the
dispersion plot: as the cap rises, more genes collapse onto the identical imputed
value, so its SD *falls* and stops representing uncertainty.

`mle` is the score to use. It fits the exponential tilt `f(x) ~ exp(theta * x)` by
censored maximum likelihood: detected genes contribute their exact position, and
the rest contribute only the probability of falling below the cap. The cap enters
the likelihood instead of the estimate, so nothing is rescaled, the mean holds in
both directions, and the spread widens honestly.

It also reports `mle_se`, from the curvature of the censored likelihood carried
onto the 0-1 scale by the delta method. **That removes the need for a separate
"undetermined" flag.** Once the signature is fully censored the likelihood
flattens, `theta` runs toward the edge of its range, and the standard error blows
up by orders of magnitude: the point estimate still says "maximally down" while
the error bar says the number carries no information. Anything consuming the score
should weight by `1 / mle_se^2` rather than test a boundary condition.

One caveat stands. In the mixture scenarios *every* estimator drifts, because
raising the cap preferentially censors whichever component sits lower and the
detected set becomes enriched for the survivor. That is a property of censoring,
not of any normaliser, so it cannot be fixed by a better denominator -- only by
modelling the two components explicitly.

```{r invariance-pipeline, include = FALSE}
gene_size <- 10000L
n_rep <- 40
rank_caps <- seq(500, 9500, by = 200)

grid <- expand.grid(
  rep = seq_len(n_rep),
  direction = c("up", "down", "flat"),
  responder_fraction = c(1, 0.3),
  stringsAsFactors = FALSE
)
# A partially-responding flat signature is still flat, so keep only one copy.
grid <- grid[!(grid$direction == "flat" & grid$responder_fraction != 1), ]
grid$scenario <- paste0(grid$direction, " (", grid$responder_fraction * 100, "% responders)")

pipe <- tt_initialise(
  store = "tidytargets-rank-cap-sweep",
  computing_resources = crew_controller_local(
    workers = max(1L, parallel::detectCores() - 1L)
  ),
  verbosity = "silent"
) |>
  tt_data(simulate_peaks) |>
  tt_data(score_variants) |>
  tt_data(score_one) |>
  tt_data(rank_caps) |>
  tt_data_list(settings <- grid |> group_split(row_number())) |>
  tt_iterate(
    sim <- score_one(settings, rank_caps, simulate_peaks, score_variants)
  ) |>
  tt_merge(sweep <- bind_rows(sim))

invisible(tt_evaluate(pipe))
sweep <- tt_explore(pipe, sweep)
```

## Scenario 1 — Signature UP (mode at 1)

### Big cell (rank_cap = 5000)

```{r up-big-cell}
set.seed(1)
plot_peak_samples(direction = "up", rank_cap = 5000)
```

### Small cell (rank_cap = 7500)

```{r up-small-cell}
set.seed(1)
plot_peak_samples(direction = "up", rank_cap = 7500)
```

### Rank-cap invariance

```{r up-invariance, fig.height = 5}
plot_invariance(sweep, "up (100% responders)")
plot_dispersion(sweep, "up (100% responders)")
gradient_table(sweep, "up (100% responders)")
```

Point estimate with its likelihood-based standard error:

```{r up-invariance-se, fig.height = 3.2}
plot_se_calibration(sweep, "up (100% responders)")
se_calibration(sweep, "up (100% responders)")
```

## Scenario 2 — Signature DOWN (mode at 0)

### Big cell (rank_cap = 5000)

```{r down-big-cell}
set.seed(1)
plot_peak_samples(direction = "down", rank_cap = 5000)
```

### Small cell (rank_cap = 7500)

```{r down-small-cell}
set.seed(1)
plot_peak_samples(direction = "down", rank_cap = 7500)
```

### Rank-cap invariance

The score must fall to 0 once the whole signature is tied below the cap. What it
must not do is climb as the cap drops, which would be depth inflating a signature
that is genuinely off.

```{r down-invariance, fig.height = 5}
plot_invariance(sweep, "down (100% responders)")
plot_dispersion(sweep, "down (100% responders)")
gradient_table(sweep, "down (100% responders)")
```

Point estimate with its likelihood-based standard error:

```{r down-invariance-se, fig.height = 3.2}
plot_se_calibration(sweep, "down (100% responders)")
se_calibration(sweep, "down (100% responders)")
```

## Scenario 3 — Signature random (flat)

### Big cell (rank_cap = 5000)

```{r random-big-cell}
set.seed(1)
plot_peak_samples(direction = "flat", rank_cap = 5000)
```

### Small cell (rank_cap = 7500)

```{r random-small-cell}
set.seed(1)
plot_peak_samples(direction = "flat", rank_cap = 7500)
```

### Rank-cap invariance

```{r random-invariance, fig.height = 5}
plot_invariance(sweep, "flat (100% responders)")
plot_dispersion(sweep, "flat (100% responders)")
gradient_table(sweep, "flat (100% responders)")
```

Point estimate with its likelihood-based standard error:

```{r random-invariance-se, fig.height = 3.2}
plot_se_calibration(sweep, "flat (100% responders)")
se_calibration(sweep, "flat (100% responders)")
```

## Scenario 4 — Partial responders (30% of the signature reacts)

Only 30% of the signature genes follow the direction; the remaining 70% sit at
background. This is the case that a one-parameter monotone family cannot
represent, so it is the fair test of the censored MLE.

### UP, 30% responders

```{r mixed-up-cell}
set.seed(1)
plot_peak_samples(direction = "up", rank_cap = 5000, responder_fraction = 0.3)
```

```{r mixed-up-invariance, fig.height = 5}
plot_invariance(sweep, "up (30% responders)")
plot_dispersion(sweep, "up (30% responders)")
gradient_table(sweep, "up (30% responders)")
```

Point estimate with its likelihood-based standard error:

```{r mixed-up-invariance-se, fig.height = 3.2}
plot_se_calibration(sweep, "up (30% responders)")
se_calibration(sweep, "up (30% responders)")
```

### DOWN, 30% responders

```{r mixed-down-cell}
set.seed(1)
plot_peak_samples(direction = "down", rank_cap = 5000, responder_fraction = 0.3)
```

```{r mixed-down-invariance, fig.height = 5}
plot_invariance(sweep, "down (30% responders)")
plot_dispersion(sweep, "down (30% responders)")
gradient_table(sweep, "down (30% responders)")
```

Point estimate with its likelihood-based standard error:

```{r mixed-down-invariance-se, fig.height = 3.2}
plot_se_calibration(sweep, "down (30% responders)")
se_calibration(sweep, "down (30% responders)")
```

# What the MLE assumes, and when that matters

The fit treats the signature's ranks as independent draws from `f(x) ~ exp(theta * x)`.
That is three assumptions: a specific family, a **monotone** density (stronger than
unimodal -- the mode is always at rank 1 or rank `G`, so an interior mode cannot be
represented at all), and independence, when real ranks are a permutation subset and
therefore mildly negatively dependent.

## The score is a moment matcher, not a shape claim

The tilt is an exponential family whose sufficient statistic is `x` itself, so its
score equation is `sum(x_i) = k * mu(theta)`. With nothing censored the fitted mean
is *identically* the sample mean, whatever the true shape:

```{r assumption-uncensored}
uncensored <- do.call(rbind, lapply(names(awkward_shapes), function(nm) {
  set.seed(31)
  peaks <- awkward_shapes[[nm]]()
  data.frame(
    shape = nm,
    mle_at_cap_0 = score_variants(peaks, 0L)[["mle"]],
    plain_mean_rank = mean((peaks - 0.5) / gene_size)
  )
}))
uncensored$difference <- uncensored$mle_at_cap_0 - uncensored$plain_mean_rank
uncensored
```

So `theta` is not a claim about the shape of the rank distribution. It is a device
for extrapolating the mean rank into the region you cannot see, and the family only
supplies a guess for the mean of the censored mass.

## The imputation is exact, not an analogy

Differentiating `log F(c; theta)` gives `E[x | x <= c] - mu(theta)`, so setting the
derivative to zero rearranges to

$$ \mu(\hat\theta) \;=\; \frac{\sum_{\text{detected}} x_i \;+\; m \cdot E[x \mid x \le c;\, \hat\theta]}{S} $$

the observed sum plus the censored count times the censored mean, over the full
signature size. The MLE's only contribution is supplying that conditional mean
self-consistently: the fitted `theta` sets the imputed value, and the imputed value
sets the fitted `theta`. It is the fixed point of the obvious EM loop.

`auc` has the same structure but plugs in `cap / 2`, the answer you get if the
hidden genes are spread uniformly below the cap. The two therefore agree for a flat
signature and diverge in proportion to how directional it is:

```{r assumption-imputation}
set.seed(1)
imputation <- do.call(rbind, lapply(c("up", "down", "flat"), function(direction) {
  peaks <- simulate_peaks(direction = direction)
  do.call(rbind, lapply(c(3000L, 5000L, 7000L), function(cap) {
    scored <- score_variants(peaks, cap)
    data.frame(
      direction, cap,
      mle_imputes = tilt$cond_mean(cap / gene_size, scored[["mle_theta"]]),
      auc_imputes = cap / gene_size / 2
    )
  }))
}))
imputation$gap <- imputation$mle_imputes - imputation$auc_imputes
imputation
```

That is also why `auc` climbed for DOWN signatures earlier. It kept imputing
`cap / 2` where the honest conditional mean was far lower, and the size of that
error grew with the cap.

## Clustered signatures: where monotonicity actually bites

These are the shapes being tested. None of them can be represented by a monotone
tilt: the mass sits in a band in the middle, or at both ends at once.

```{r assumption-shapes, fig.height = 3.6}
plot_awkward_shape("band 5000-7000", rank_cap = 1000)
plot_awkward_shape("narrow block at 6000", rank_cap = 1000)
plot_awkward_shape("two clusters, 2k and 9k", rank_cap = 1000)
```

Note the scores in those panels. The cap sits below every cluster, so all 1000
genes are detected and the `mle` is already exactly right -- 0.601, 0.600 and
0.550 against true means of 0.601, 0.600 and 0.550 -- despite the shape being
entirely wrong for the model. Nothing is censored, so the fit is pure moment
matching and the family never gets consulted.

Now move the cap into the cluster. Same signature, same everything, only the cap
has changed:

```{r assumption-shapes-censored, fig.height = 3.6}
plot_awkward_shape("band 5000-7000", rank_cap = 6500)
plot_awkward_shape("narrow block at 6000", rank_cap = 6500)
plot_awkward_shape("two clusters, 2k and 9k", rank_cap = 8000)
```

The band at 5000-7000 drops from 0.601 to 0.371, and the narrow block collapses
from 0.600 to 0.017 once its 1000 genes fall below the cap in one go. In both
cases the model imputes the hidden genes as though the density kept falling toward
0, when really they sit immediately under the cap. The two-cluster case moves the
other way, 0.550 to 0.725, because censoring removes its lower lobe and leaves the
upper one: same mechanism as the partial-responder scenarios, since a two-cluster
signature *is* a mixture of an UP and a DOWN population.

The full sweep, with the diagnostics:

```{r assumption-clusters}
clusters <- do.call(rbind, lapply(names(awkward_shapes), function(nm) {
  cbind(shape = nm, shape_table(nm, caps = c(3000L, 5000L, 6500L, 8000L)))
}))
clusters
```

Two things to read off. First, `bias` stays at machine zero until `detected` starts
falling, then grows fast, exactly as the pairs of figures above show.

Second, and importantly, **`se` does not react**. A standard error measures sampling
noise *under the assumed model* and is structurally blind to the model being wrong.
Everything said earlier about the SE flagging trouble applies to running out of
data, not to misspecification.

## Two diagnostics that do react

`ks` is the largest gap between the detected ranks and the fitted tilt, compared
against a 5% critical value near `1.36 / sqrt(k)`, about 0.043 at 1000 genes. Read
it as a magnitude rather than a verdict. The monotone signature sits just above the
line at 0.05-0.07, because `Beta(5, 1)` is not literally an exponential tilt either
-- only mildly misspecified. The banded and two-cluster signatures land at 0.31 to
0.82, an order of magnitude further out, which is the regime where the imputation
cannot be trusted.

`bound` needs no assumption about shape at all. Both the score and the truth have
the form `(sum_detected + m * E) / S` and differ only in `E`, which must lie in
`[0, cap]`. So the error cannot exceed the censored fraction times the cap:

```{r assumption-bound}
binding <- clusters[clusters$bound > 0, ]
data.frame(
  cases = nrow(clusters),
  bound_holds = sum(abs(clusters$bias) <= clusters$bound + 1e-8),
  # Rows with no censoring have a bound of exactly 0, so the ratio is only
  # meaningful where the bound actually binds.
  worst_ratio_where_censored = max(abs(binding$bias) / binding$bound)
)
```

That gives the operating rule. When `bias_bound` is small the score is trustworthy
whatever the true shape; when it is large the answer rests entirely on the shape
assumption, and `ks` tells you whether that assumption is holding.

## Independence

Ranks are drawn without replacement, so the i.i.d. likelihood slightly overstates
the spread. The finite-population correction is `sqrt(1 - S/G)`, about 0.95 here, so
the effect is a few percent in the conservative direction. It is part of why the
observed `se_over_sd` ratios above sit a little over 1 rather than under, and it is
not worth correcting.
